metric analysis

### 2021-05-30 metric_analysis.Rmd
Analyze both resistance and resilience

@@

library(readr)
resistance <- read_csv("ED-outputs/results/metric/resistance_values.csv")
resilience <- read_csv("ED-outputs/results/metric/resilience_values.csv")
met <- read_csv("ED-outputs/results/constant_annual_met_data.csv")
```

@@

# Met-metric data

@@

```{r, fig.height=8}
# Build a combined met-metric dataset
resistance %>% 
  select(variable, met, trough_resistance) %>% 
  # We see from the met analysis above that vbdsf and vddsf are highly
  # correlated with nbdsf and nddsf respectively; remove one pair
  left_join(select(met2_wide, -nbdsf, -nddsf), by = "met") ->
  resist_combined

resist_combined %>% 
  pivot_longer(dlwrf:vgrd) %>% 
  ggplot(aes(value, trough_resistance, color = variable)) +
  geom_point() + 
  facet_wrap(variable~name, scales = "free") + 
  geom_smooth(method = "lm", formula = y ~ x, color = "black", linetype = 5) +
  theme(legend.position = "none", axis.text = element_text(size = 5))
```

# Variable importance

@@

```{r, message=FALSE}
# Fit a Random Forest for each variable (across all treatment severities)
# Note the r.f. functions don't play well with tibbles, so change to data frame
library(randomForest)
rc_all <- split(as.data.frame(resist_combined), resist_combined$variable)
rf_all <- lapply(rc_all, function(x) 
  randomForest(trough_resistance ~ ., data = x[-1:-2], importance = TRUE)
)

# Importance metrics and partial plot data
import <- list()
partial <- list()
for(var in names(rf_all)) {
  mod <- rf_all[[var]]
  imp <- as.data.frame(importance(mod))
  imp$met_var <- rownames(imp)
  import[[var]] <- as_tibble(imp)

  impvar <- rownames(imp)[order(imp[, 1], decreasing = TRUE)]
  for (i in seq_along(impvar[1:3])) {
    pp <- partialPlot(mod, rc_all[[var]], impvar[i], plot = FALSE)
    partial[[paste(var, i)]] <- tibble(var = var, 
                                       met_var = impvar[i],
                                       x = pp$x, y = pp$y)
  }
}

import <- bind_rows(import, .id = "var")
library(forcats)
lapply(split(import, import$var), function(x) {
  ggplot(x, aes(fct_reorder(met_var, `%IncMSE`), `%IncMSE`)) + 
    geom_col() +
    coord_flip() +
    xlab("Met variable") +
    ggtitle(unique(x$var))
})

partial <- bind_rows(partial)
ggplot(bind_rows(partial), aes(x, y, color = var)) + 
  geom_line() + 
  xlab("Met var value") + ylab("Metric") +
  ggtitle("Partial dependence plot") +
  facet_grid(var ~ met_var, scales = "free")
```

@@

Interestingly, `tmp` (air temperature) and `vbdsf` (sunlight) are among the
top-three most important drivers for all four output variables 
(GPP, NEP, NPP, Rh) for resistance. Atmospheric pressure (`pres`) is
among the top three for both GPP and NPP resistance: the former because of
stomatal effects perhaps (?), and the latter because
of NPP's strong dependence on GPP.

>>>>

set.seed(1234)

@@

library(readr)
resistance <- read_csv("ED-outputs/results/metric/resistance_values.csv")
resilience <- read_csv("ED-outputs/results/metric/resilience_values.csv")
met <- read_csv("ED-outputs/results/constant_annual_met_data.csv")

resistance %>% 
  select(scn, variable, met, value = trough_resistance, severity) %>% 
  mutate(metric = "resistance") ->
  combined
resilience %>% 
  select(scn, variable, met, value = resilience, severity) %>% 
  mutate(metric = "resilience") %>% 
  bind_rows(combined) ->
  combined
```

@@

# Met-metric data {.tabset}

@@

## Resistance

```{r, fig.height=8}
# Build a combined met-metric dataset
combined %>% 
  select(metric, variable, met, value) %>% 
  # We see from the met analysis above that vbdsf and vddsf are highly
  # correlated with nbdsf and nddsf respectively; remove one pair
  left_join(select(met2_wide, -nbdsf, -nddsf), by = "met") ->
  metric_combined

resistance %>% 
  select(variable, met, trough_resistance) %>% 
  left_join(select(met2_wide, -nbdsf, -nddsf), by = "met") ->
  resist_combined

metric_combined %>% 
  pivot_longer(dlwrf:vgrd, values_to = "met_value") ->
  metric_comb_long

p_mm <- ggplot(filter(metric_comb_long, metric == "resistance"), 
            aes(met_value, value, color = variable)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, color = "black", linetype = 5) +
  facet_wrap(~ variable + name, scales = "free") + 
  ggtitle("Resistance") +
  theme(legend.position = "none", axis.text = element_text(size = 5))
print(p_mm)
```

## Resilience

```{r, fig.height=8}
p_mm <- p_mm %+% filter(metric_comb_long, metric == "resilience")
print(p_mm + ggtitle("Resilience"))
```

# Variable importance {.tabset}

@@

```{r, message=FALSE}
# Fit a Random Forest for each variable (across all treatment severities)
# Note the r.f. functions don't play well with tibbles, so change to data frame
library(randomForest)
rc_all <- split(as.data.frame(metric_combined), ~ metric + variable)
rf_all <- lapply(rc_all, function(x) 
  # remove metric, variable, met columns and run RF
  randomForest(value ~ ., data = x[-1:-3], importance = TRUE)
)

# Importance metrics and partial plot data
import <- list()
partial <- list()
for(var in names(rf_all)) {
  mod <- rf_all[[var]]
  metric <- rc_all[[var]]$metric[1]
  variable <- rc_all[[var]]$variable[1]

  imp <- as.data.frame(importance(mod))
  imp$met_var <- rownames(imp)
  imp$metric <- metric
  imp$variable <- variable
  import[[var]] <- as_tibble(imp)

  impvar <- rownames(imp)[order(imp[, 1], decreasing = TRUE)]
  for (i in seq_along(impvar[1:3])) {
    pp <- partialPlot(mod, rc_all[[var]], impvar[i], plot = FALSE)
    partial[[paste(var, i)]] <- tibble(metric = metric,
                                       variable = variable,
                                       met_var = impvar[i],
                                       x = pp$x, y = pp$y)
  }
}

import <- bind_rows(import)
partial <- bind_rows(partial)
```

## Resistance

```{r}
library(forcats)
f_imp_plot <- function(x) {
  ggplot(x, aes(fct_reorder(met_var, `%IncMSE`), `%IncMSE`)) + 
    geom_col() +
    coord_flip() +
    xlab("Met variable") +
    ggtitle(unique(x$variable))
}
lapply(split(filter(import, metric == "resistance"), ~ variable), f_imp_plot)

p_pdp <- ggplot(filter(partial, metric == "resistance"), 
                aes(x, y, color = variable)) + 
  geom_line() + 
  xlab("Met var value") + ylab("Resistance") +
  ggtitle("Partial dependence plot") +
  facet_grid(variable ~ met_var, scales = "free")
print(p_pdp)
```

## Resilience

```{r}
# Re-do above plots but now for resilience
lapply(split(filter(import, metric == "resilience"), ~ variable), f_imp_plot)
p_pdp <- p_pdp %+% filter(partial, metric == "resilience")
print(p_pdp + ylab("Resilience"))
```

@@

For **resistance**, `tmp` (air temperature) and `vbdsf` (sunlight) are among the
top-three most important drivers for all four output variables 
(GPP, NEP, NPP, Rh). Higher temperature and more light translate into higher
resistance for all variables.

This is largely true for **resilience** as well, although
`prate` (precipitation) is important for both NPP and Rh; in fact precipitation
is far and away the most important variable for Rh. Here, however,
higher temperature and more light mean _less_ resilience.
